---
title: "Spatial Data Analysis in R: `raster` and `terra`"
description: |
  `raster` has been the go-to spatial analysis package in R since the dawn of time, but it's time to convert our workflow to the more modern `terra` package. At the Ocean Health Index, we work with spatial data for a plethora of goals like mapping sea ice extent and habitat types that sequester carbon throughout the land and coastal nations. As we map the 2022 sea level rise across the globe, let's dive into how we convert our `raster` functions into `terra` functions.
author:
  - name: Juliet Cohen
    url: {}
date: 2022-06-16
output:
  distill::distill_article:
    self_contained: false
draft: TRUE
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

Spatial data in R has a reputation for being tedious and time consuming. It can be challenging to wrangle `.shp` files, `.nc` files, and `.tif` files and their metadata in order to execute analyses and visualize rasters. The [Ocean Health Index](https://oceanhealthindex.org/) has historically applied the `raster` package to annually monitor the relationship between the marine health and human well-being for 220 regions. But with the goal of continuously improving methodology and keeping up with the hip trends in environmental science, in 2022 the Ocean Health Index began calculating `sea ice extent`, `habitat mapping`, and `sea level rise` using the `terra` package instead of the `raster` package. 

##### How do we go about converting spatial functions between these packages and executing intermediate checks along the way?

### 1. Sea Level Rise

Let's start with the basics: Import a `.tif` file that represents the world's ocean as a raster using `raster::raster()` versus `terra::rast()` and plot both to compare the output.

```{r, eval=FALSE}
ocean_raster <- raster::raster(file.path(dir_M, "git-annex/globalprep/spatial/v2017/ocean.tif"))
plot(ocean_raster)

ocean_terra <- terra::rast(file.path(dir_M, "git-annex/globalprep/spatial/v2017/ocean.tif"))
plot(ocean_terra)
```

They sure do look similar. Now let's use the grid cell values of these spatial objects to populate the cells of a dataframe, which allows us to compare the values at each latitude and longitude pairing between packages. Raster documentation shows that we can use the `as.data.frame()` from base R for both packages. 

```{r}
### Idea for this script: bring in the monthly file we used in the OHI script because it is more explanatory

# convert to df's
ocean_raster_df <- as.data.frame(ocean_raster_agg, xy = TRUE, na.rm = FALSE)
ocean_terra_df <- as.data.frame(ocean_terra_agg, xy = TRUE, na.rm = FALSE)

# if ncessary, recode the NA value for the spatRaster we made with `terra` 


# compare the two rasterLayer to spatraster
# set the two dataframes equal to each other to populate the cells of a new dataframe
# cells with TRUE are those which matched in both dataframes, and cells with FALSE are those which did not match
true_false_df <- as.data.frame(raster_y2020_m03_df == terra_y2020_m03_df)
# all values are NA or TRUE, no FALSE values in any column

# count the NA values in each df, they should be the same
# count number of NA values in each df, they are the same!
sum(is.na(raster_y2020_m03_df$Averaged.Sea.Level.Anomalies.2020.03)) # 433059
sum(is.na(terra_y2020_m03_df$sla)) # 43059

# last check: subtract one df's values from the other

```

Casey's workflow for reading in .tif file and converting to raster & df:

```{r}
ocean_10km_file <- here('_spatial', 'ocean_area_mol.tif')
# unlink(ocean_10km_file)

if(!file.exists(ocean_10km_file)) {
  message('Processing ocean area at 10 km resolution...')
  ocean_moll_1km <- raster(ocean_1km_file)
  
  ocean_moll_10km <- aggregate(ocean_moll_1km, fact = 10, fun = sum, progress = 'text')
  
  # rast_base_10km <- raster(rast_base_10km_file)
  # compareRaster(ocean_moll_10km, rast_base_10km)
  # [1] TRUE
  
  ### drop the Caspian Sea:
  casp_bbox <- c(xmin = 3700000, xmax = 4800000, ymin = 4300000, ymax = 5650000)
  ocean_moll_df <- as.data.frame(ocean_moll_10km, xy = TRUE) %>%
    setNames(c('x', 'y', 'z')) %>%
    mutate(caspian_x = between(x, casp_bbox['xmin'], casp_bbox['xmax']),
           caspian_y = between(y, casp_bbox['ymin'], casp_bbox['ymax']),
           caspian   = caspian_x & caspian_y) %>%
    mutate(z = ifelse(caspian, NA, z)) %>%
    select(-contains('caspian'))
  
  ocean_moll_10km_no_casp <- rasterFromXYZ(ocean_moll_df, res = 10000, crs = crs(ocean_moll_10km))
  
  # plot(ocean_moll_10km_no_casp)
  ocean_moll_10km_no_casp <- ocean_moll_10km_no_casp / 100
  
  writeRaster(ocean_moll_10km_no_casp, ocean_10km_file, overwrite = TRUE)
}
```

### Raster Manipulation with `aggregate()`

We can use the `aggregate()` function to lower the spatial resolution by a factor of 10:
- `fact` argument represents the aggregation factor (the number of cells in each direction horizontally and vertically), a value of 10 results in a raster with 10 kilometer resolution if the original `.tif` file is at 1 kilometer resolution
- `fun` argument is the function used to aggregate the values
- `progress` argument shows a progress bar as the raster is aggregated, this is helpful for large raster files (when executing in a chunk rather than a script)

```{r}
ocean_raster_agg <- raster::aggregate(ocean_raster, fact = 10, fun = sum, progress = 'text')
ocean_terra_agg <- terra::aggregate(ocean_terra, fact = 10, fun = sum, progress = 'text')
```














