---
title: "A function I learned"
description: |
  A short description of the post.
author:
  - name: Unknown
    url: {}
date: 08-10-2021
output:
  distill::distill_article:
    self_contained: false
preview: Images/bembe.jpg
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(palmerpenguins)
```

## abline function

Given a graph of penguin body mass as a function of penguin flipper length, the geom_smooth(method = lm) function creates a best fit line for the scatterplot.

### Within the function geom_smooth(), you can specify if you want the line to show a confidence interval with se = FALSE or TRUE.

### Furthermore, you can add linetype = "dashed" to change the line type.

```{r}
ggplot(data = penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point() +
  labs( x = "Flipper length (mm)", 
        y = "Body mass (g)", 
        main = "Penguin Body Mass as a Function of Flipper Length") +
  geom_smooth(method = lm, se = FALSE, linetype = "dashed")

```



Distill is a publication format for scientific and technical writing, native to the web.

Learn more about using Distill at <https://rstudio.github.io/distill>.


